---
title: "toothGrowth"
author: "Karim Qazi"
date: "September 19, 2014"
output: html_document
---
**Load the ToothGrowth data and perform some basic exploratory data analyses:**
```{r echo=TRUE}
library(knitr)
data(ToothGrowth)
head(ToothGrowth)
summary(ToothGrowth)
vitaminCGrowth <- ToothGrowth[which(ToothGrowth$supp == "VC"),]
ojGrowth <- ToothGrowth[which(ToothGrowth$supp == "OJ"),]
```

**Mean of Vitamin C tooth growth by dosage:**
```{r echo=TRUE}
meanVCGrowthByDose <- aggregate(x = vitaminCGrowth$len, by = list(vitaminCGrowth$dose), mean)
colnames(meanVCGrowthByDose)[1:2] <- c("Dosage", "Tooth Growth")
kable(meanVCGrowthByDose, format = "pandoc")
```

**Mean of OJ tooth growth by dosage:**
```{r echo=TRUE}
meanOJGrowthByDose <- aggregate(x = ojGrowth$len, by = list(ojGrowth$dose), mean)
colnames(meanOJGrowthByDose)[1:2] <- c("Dosage", "Tooth Growth")
kable(meanOJGrowthByDose, format = "pandoc")
```

Execute T-test with "paired" since the same 10 guinea pigs were given three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid) (per TootGrowth docs).  Also with such low sample sizes (10 each) a T-test would be most applicable here.

**Confidence interval for tooth growth differnces between OJ and Vitamin C:**

```{r echo=TRUE}
lowDosageTest <- t.test(ojGrowth[which(ojGrowth$dose == 0.5),]$len, 
                        vitaminCGrowth[which(vitaminCGrowth$dose == 0.5),]$len, 
                        paired=TRUE)
midDosageTest <- t.test(ojGrowth[which(ojGrowth$dose == 1.0),]$len, 
                        vitaminCGrowth[which(vitaminCGrowth$dose == 1.0),]$len, 
                        paired=TRUE)
highDosageTest <- t.test(ojGrowth[which(ojGrowth$dose == 2.0),]$len, 
                         vitaminCGrowth[which(vitaminCGrowth$dose == 2.0),]$len, 
                         paired=TRUE)
```
```{r echo=FALSE}
dosageResults <- data.frame(c("Low Dosage", "Mid Dosage", "High Dosage"), 
                            c(paste(lowDosageTest$conf.int[1:2], collapse = ' '), 
                              paste(midDosageTest$conf.int[1:2], collapse = ' '), 
                              paste(highDosageTest$conf.int[1:2], collapse = ' ')), 
                            c(lowDosageTest$p.value, midDosageTest$p.value,highDosageTest$p.value))
colnames(dosageResults)[1:3] <- c("Dosage", "Paired t-test confidence interval", "P value")
```

Below are the various confidence intervals and p-values for each dosage when comparing OJ to Vitamin C as it relates to tooth growth:
```{r echo=FALSE}
kable(dosageResults, format = "pandoc")
```
**Conclusions**

The hypothesis test is below, where the value of **mu** is the mean of Vitamin C tooth growth for the low and mid dosage, so rejecting the NULL hypothesis would indicate that OJ leads to greather tooth growth than Vitamin C and vice versa.  The reason for only evaluating the **low** and **mid** dosage is because these had p-values that have a high probability of rejecting the NULL hypothesis.

Assuming a low dosage p-value of **`r lowDosageTest$p.value`** and mid dosage p-value of **`r midDosageTest$p.value`** we can **reject** the NULL hypotheseis and conclude that OJ leads to greather tooth growth than Vitamin C for these dosage levels, assuming a 0.05 level of signifigance (or 95% confidence level) .

**Low Hypothesis Test**

H0 = mu = **`r meanVCGrowthByDose[1,2]`**

HA = mu > **`r meanVCGrowthByDose[1,2]`**

**Mid Hypothesis Test**

H0 = mu = **`r meanVCGrowthByDose[2,2]`**

HA = mu > **`r meanVCGrowthByDose[2,2]`**