---
title: "regression-analysis.Rmd"
author: "Karim Qazi"
date: "October 26, 2014"
output: pdf_document
---
**Executive Summary**

We are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome).  This report attempts to answer the following questions as it relates to this high level goal:

1. Is an automatic or manual transmission better for MPG?
2. Quantify the MPG difference between automatic and manual transmissions?

We will use the **mtcars** data set to perform this analysis.  Both exploratory and inferential analsysis were conducted to create the results below.

*Conclusions*

The following factors illustrate why **weight** is a much better predictor of **mpg** than **transmission** is:

1. Looking at *Figure 1* in the appendix it's clear that this doesn't fit a linear model well. So I wouldn't use **transmission** to predict **mpg**. 
2. Evaluating all the coefficents for potential predictors of **mpg**, **weight** looks to do a good job of this with ~**3.72** decrease in **mpg**.
3. The low **p-value** when using **weight** to predict **mpg** also supports a good probability for this.
3. Looking at *Figure 2* in the appendix it's clear that **weight**  IS a good predictor of **mpg**.
4. The very low residual of **weight** to predict **mpg** provides futher evidence of this.  See *Figure 3* in the appendix to see this plotted as well as the actual residual value in the analysis below.

```{r echo=FALSE}
library(knitr)
data(mtcars)
```
**Exploratory Data Analysis**

```{r echo=FALSE}
avgTransMPG <- aggregate(x = mtcars$mpg, by=list(mtcars$am), mean)
colnames(avgTransMPG)[1:2] <- c("Transmission", "MPG")
```
We can see that on average Manual transmission cars get ~ **`r round(avgTransMPG$MPG[1] / avgTransMPG$MPG[2], 3) * 100`** better MPG when compared to Automatic transmissions.  Which looks pretty compelling, but we haven't done enough analysis to quantify this just yet.
```{r echo=FALSE}
kable(avgTransMPG)
```

Another logical predictor of **mpg** could be **weight**.  Looking at weight we can see it looks to be a potentially better predictor of MPG than transmission type is.  Weight is the estimated expected change in **mpg** per 1,000 lb increase in weight.

**NOTE:** I rounded up the weight in tons so you could get a sense of this more easily, but this does loose valuable precision that must be accounted for in later analysis.
```{r echo=FALSE}
avgWeightMPG <- aggregate(x = mtcars$mpg, by=list(round(mtcars$wt, 0)), mean)
colnames(avgWeightMPG)[1:2] <- c("Weight (per 1,000 lb)", "MPG")
kable(avgWeightMPG)
```

This analysis will provide some guidance for what's to follow below.

**Interpretation of Coefficents**

Let's look at all the variables in the data set we can potentially predict **mpg** to all the other potential predictors of **mpg**.  Since we originally wanted to see if **transmission** could predict **mpg** we can do a summary of all the coefficents.  Seeing that **am** (which indicates transmission) looks to predict ~**2.52** increase in **mpg**, whereas **weight** (per 1,000 lb) predicts a decrease of **~3.72** leads us to believe that **weight** is  a better predictor of **mpg** from this analysis.
```{r echo=TRUE}
summary(lm(mpg ~ . , data=mtcars))$coef
```

Next let's see how we'll this **weight** predictor is performing from a probability standpoint.  If we look at the "p-value" when using weight as a predictor of "mpg" we can see that it has a very low p-value of **1.293959e-10** which should indicate that weight has a very good probability of predicting "weight".
```{r echo=TRUE}
x <- mtcars$wt
y <- mtcars$mpg
n <- length(y)
fit <- lm(y ~ x)
summary(fit)$coef
```

All the basic data points are still supporting that **weight** is a good predictor of **mpg** with ~`r summary(fit)$coef[2,1]` decrease in **mpg**.  Let's plot the data and see how it looks in **Figure 2** in the appendix. **weight** looks to predict **MPG** well from a linear regression standpoint.

To validate that this is truly a good fit we can look at the residual variation between the actual value and the predictor when taking into account **weight** to predict **mpg**.  You can see that the **max** residual is very, very small below.  You can see this plotted by looking at **Figure 3** in the appendix.
```{r echo=TRUE}
e <- resid(fit)
yhat <- predict(fit)
max(abs(e -(y - yhat)))
```

**Appendix**

***Figure 1***

Illustrates that using transmission type to predict MPG is a poor model fit.
```{r}
plot(mtcars$am, mtcars$mpg, xlab="Transmission (0=Automatic, 1=Manual)", ylab="MPG")
abline(lm(mtcars$mpg ~ mtcars$am))
```

***Figure 2***

Linear fit of when using car weight to predict MPG.

```{r}
plot(mtcars$wt, mtcars$mpg, xlab="Weight (per 1,000 lbs)", ylab="MPG")
abline(lm(mtcars$mpg ~ mtcars$wt))
```

***Figure 3***

Residual variation for linear model using weight to predict MPG.
```{r}
plot(x, y,  
     xlab = "Car weigth (per 1,000 lb)", 
     ylab = "MPG", 
     bg = "lightblue", 
     col = "black", cex = 2, pch = 21,frame = FALSE)
abline(fit, lwd = 2)
for (i in 1 : n) 
  lines(c(x[i], x[i]), c(y[i], yhat[i]), col = "red" , lwd = 2)
```


